---
format:
  pdf:
    keep-tex: true
    documentclass: scrartcl
    title-block-style: default
    crossref:
      tbl-prefix: "Table S"
    include-in-header:
      - text: |
          \usepackage{pdflscape}
          \usepackage{pifont}
          \usepackage{booktabs}
          \usepackage{tabularx}
          \usepackage{array}
          \usepackage{ltablex}
          \keepXColumns
          \setlength{\LTleft}{0pt}
          \setlength{\LTright}{0pt}
          \newenvironment{abstractsmall}{\begin{small}}{\end{small}}
          \BeforeBeginEnvironment{abstract}{\begin{abstractsmall}}
          \AfterEndEnvironment{abstract}{\end{abstractsmall}}
          % Supplemental numbering: Table S1, S2, ...
          \renewcommand{\thetable}{S\arabic{table}}
          \renewcommand{\thefigure}{S\arabic{figure}}
editor: visual
bibliography: "D:/varios/monitor_congreso/pape/polarizacion/docs/references_polarization.bib"

execute:
  echo: false
  message: false
  warning: false
---

\begin{center}
{\LARGE Supplemental materials for "When Democratic Institutions Polarize: Evidence from Legislative Behavior in Chile"\par}
\vspace{0.75em}
\end{center}
\vspace{1.5em}

## SM1: Dimensionality

A key modeling choice in the main text is to estimate dynamic ideal points in one latent dimension. This choice is not merely conventional: the substantive quantities of interest (distance-based measures of extremity and chamber-level dispersion over time) are meaningful only relative to the geometry of the underlying preference space. If roll-call voting were genuinely multi-dimensional in a stable way, a one-dimensional summary could distort distances and misstate polarization trends. For that reason, it is standard in roll-call research to treat one-dimensional dynamics as the baseline only after checking whether an additional dimension contributes explanatory power [@clinton_jackman_rivers_2004_rollcall, @poole_rosenthal_1997_congress].

We assess dimensionality using a two-dimensional DW-NOMINATE implementation and extract eigenvalues for each period’s roll-call vote matrix. Each estimation uses either the right-wing or the left-wing deputy mentioned in the main text as the anchor. Table S1 reports the leading eigenvalues $\lambda_1,\lambda_2$, their ratio $\lambda_1/\lambda_2$, and implied variance shares $\lambda_k/(\lambda_1+\lambda_2)$ by period. When $\lambda_1/\lambda_2$ is large and the first-dimension share is high, the empirical structure of voting is predominantly one-dimensional, and focusing the dynamic analysis on the first dimension is appropriate for measuring polarization along the main cleavage.

In this dataset, across all periods, $\lambda_1$ is at least fourteen times larger than $\lambda_2$, implying that the first dimension accounts for more than 93% of the overall dispersion in the roll-call data within each period. Given the dominance of the first dimension, the added complexity of a dynamic two-dimensional specification is unlikely to change inference about the main polarization trend, while potentially reducing interpretability. Finally, anchor choice affects only the orientation (sign) of the recovered scale; it does not affect eigenvalue-based dimensionality diagnostics, which are invariant to sign reversals.

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='asis'}
## S1. Dimensionality diagnostics (DW-NOMINATE 2D)

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(stringr)
  library(knitr)
  library(here)
})

aqui <- here()

path_m1 <- paste0(aqui, "/pape/polarizacion/output/wnominate_dim_eigen_by_period.csv")
eig <- read_csv(path_m1, show_col_types = FALSE)

pick_first <- function(df, pattern) {
  nm <- names(df)
  hit <- nm[str_detect(tolower(nm), pattern)]
  if (length(hit) == 0) NA_character_ else hit[1]
}

# --- Column detection (robust) ---
col_period    <- pick_first(eig, "period(_num)?|legisl|term|session|period_label|periodo")
col_l1        <- pick_first(eig, "^(lambda1|eig1|eigen1|l1)$|lambda_?1|eig_?1|eigen_?1")
col_l2        <- pick_first(eig, "^(lambda2|eig2|eigen2|l2)$|lambda_?2|eig_?2|eigen_?2")
col_ratio     <- pick_first(eig, "ratio.*1.*2|lambda1_2|l1_2")
col_anchor_id <- pick_first(eig, "anchor_id|ancla|anchor")
col_polar_id  <- pick_first(eig, "polarity_id|polarity")
col_fallback  <- pick_first(eig, "use_fallback|fallback")

if (is.na(col_l1) || is.na(col_l2)) {
  stop("Could not identify the eigenvalue columns for dimensions 1 and 2 in wnominate_dim_eigen_by_period.csv.")
}
if (is.na(col_anchor_id) || is.na(col_polar_id) || is.na(col_fallback)) {
  stop("Could not identify anchor_id / polarity_id / use_fallback columns in wnominate_dim_eigen_by_period.csv.")
}

# Anchor IDs (as specified)
RIGHT_ID <- 917
LEFT_ID  <- 837
CHECK <- "\\ding{51}"

eig2 <- eig %>%
  mutate(
    period_id = if (!is.na(col_period)) as.character(.data[[col_period]]) else as.character(row_number()),
    lambda1   = as.numeric(.data[[col_l1]]),
    lambda2   = as.numeric(.data[[col_l2]]),
    ratio_1_2 = if (!is.na(col_ratio)) as.numeric(.data[[col_ratio]]) else lambda1 / lambda2,
    share_dim1 = lambda1 / (lambda1 + lambda2),
    share_dim2 = lambda2 / (lambda1 + lambda2),
    anchor_used = if_else(
      as.logical(.data[[col_fallback]]),
      as.integer(.data[[col_polar_id]]),
      as.integer(.data[[col_anchor_id]])
    )
  )

tab_s1 <- eig2 %>%
  group_by(period_id) %>%
  summarise(
    lambda1    = first(lambda1),
    lambda2    = first(lambda2),
    ratio_1_2  = first(ratio_1_2),
    share_dim1 = first(share_dim1),
    share_dim2 = first(share_dim2),
    `right-wing anchor` = if_else(any(anchor_used == RIGHT_ID, na.rm = TRUE), CHECK, ""),
    `left-wing anchor`  = if_else(any(anchor_used == LEFT_ID,  na.rm = TRUE), CHECK, ""),
    .groups = "drop"
  ) %>%
  mutate(period_num = suppressWarnings(as.integer(period_id))) %>%
  arrange(period_num, period_id) %>%
  select(-period_num)

col_names <- c(
  "period",
  "$\\lambda_1$",
  "$\\lambda_2$",
  "$\\lambda_1/\\lambda_2$",
  "share dim 1",
  "share dim 2",
  "right-wing anchor",
  "left-wing anchor"
)

align_vec <- c("l", "c", "c", "c", "c", "c", "c", "c")

kable(
  tab_s1,
  format = "latex",
  booktabs = TRUE,
  digits  = 3,
  col.names = col_names,
  align   = align_vec,
  escape  = FALSE,
  caption = "DW-NOMINATE 2D dimensionality diagnostics by period: eigenvalues, dominance ratio ($\\lambda_1/\\lambda_2$), implied variance shares, and anchors used."
)


```

## SM2: Robustness to Non-Voting (Absences and Abstentions)

Non-voting (absences and abstentions) can in principle distort ideal-point estimation if it is systematically related to legislators’ latent preferences or changes over time. We therefore re-estimate the baseline dynamic IRT model (three-anchor specification) after excluding legislators whose non-vote rate within any served period exceeds alternative thresholds (25%, 35%, and 50%), where “non-vote” is defined as any response other than Yea ("Afirmativo") or Nay ("En Contra"}. Numbers and proportions are in Table S2. The descriptive diagnostics show that absences decline over time. The robustness results in Table S3 indicate that filtering extreme non-voters leaves both individual ideal points and the period-level polarization series essentially unchanged: correlations with the baseline estimates are uniformly near 1, and regressions of filtered estimates on baseline estimates yield slopes close to 1 and intercepts close to 0. These results imply that the main polarization trend is not an artifact of participation patterns or a small number of low-participation legislators.

```{r m2_nonvotes_summary_tables, echo=FALSE, message=FALSE, warning=FALSE}

# ============================================================
# M2 — Robustness to non-votes (absences + abstentions)
# Reads pre-computed outputs (no re-estimation):
#   - summary_nonvote_sensitivity.csv   (comparisons vs baseline)
#   - exclusion_counts.csv              (# excluded by threshold)
# Produces:
#   - Table S4: Exclusion counts by threshold
#   - Table S5: Stability diagnostics (correlations + regressions) vs baseline
# ============================================================

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(knitr)
  library(here)
})

aqui  <- here()
dir_m2 <- file.path(aqui, "pape", "polarizacion", "output", "sensitivity_nonvotes")

path_sum  <- file.path(dir_m2, "summary_nonvote_sensitivity.csv")
path_excl <- file.path(dir_m2, "exclusion_counts.csv")

stopifnot(file.exists(path_sum), file.exists(path_excl))

sumdf  <- read_csv(path_sum,  show_col_types = FALSE)
excldf <- read_csv(path_excl, show_col_types = FALSE)

# --- Helper: consistent threshold labels --------------------------------------
pretty_thr <- function(x) {
  dplyr::case_when(
    x %in% c("t25","filtered_t25") ~ "25%",
    x %in% c("t35","filtered_t35") ~ "35%",
    x %in% c("t50","filtered_t50") ~ "50%",
    x %in% c("full","filtered_full") ~ "Full (baseline)",
    TRUE ~ as.character(x)
  )
}

# --- Helper: force exact columns (adds missing as NA, drops extras) -----------
ensure_cols <- function(df, cols) {
  for (cl in cols) if (!cl %in% names(df)) df[[cl]] <- NA
  dplyr::select(df, dplyr::all_of(cols))
}

# -------------------- Table S4: Exclusions ------------------------------------
cols_s4 <- c("threshold","excluded_n","excluded_pct","kept_n","total_n")

tab_s4 <- if (all(cols_s4 %in% names(sumdf))) {
  sumdf %>%
    distinct(threshold, excluded_n, excluded_pct, kept_n, total_n)
} else {
  # fallback: older exclusion_counts.csv only has counts
  # try to detect the count column robustly
  count_col <- if ("n_excluded" %in% names(excldf)) "n_excluded" else names(excldf)[2]
  excldf %>%
    transmute(
      threshold   = .data[["threshold"]],
      excluded_n  = as.integer(.data[[count_col]]),
      excluded_pct = NA_real_,
      kept_n      = NA_integer_,
      total_n     = NA_integer_
    )
}

tab_s4_print <- ensure_cols(tab_s4, cols_s4) %>%
  mutate(
    threshold = pretty_thr(threshold),
    excluded_pct = ifelse(!is.na(excluded_pct) & excluded_pct <= 1, 100 * excluded_pct, excluded_pct)
  ) %>%
  arrange(match(threshold, c("Full (baseline)","25%","35%","50%")), threshold)

kable(
  tab_s4_print,
  digits = 2,
  col.names = c("Threshold", "Excluded (n)", "Excluded (%)", "Kept (n)", "Universe (n)"),
  caption = "Legislator exclusions under alternative non-vote thresholds (non-vote = any response other than {Afirmativo, En Contra})."
)

# -------------------- Table S5: Stability diagnostics -------------------------
cols_s5 <- c("threshold","variable_type","n","pearson","spearman","reg_intercept","reg_slope","reg_r2")
stopifnot(all(cols_s5 %in% names(sumdf)))

tab_s5_print <- sumdf %>%
  filter(threshold != "full") %>%   # comparisons defined vs baseline
  mutate(
    threshold = pretty_thr(threshold),
    variable_type = dplyr::recode(
      variable_type,
      ideology_points = "Ideal points (DiputadoId × Periodo)",
      polarization_series = "Polarization series (period means)"
    )
  ) %>%
  ensure_cols(cols_s5) %>%
  arrange(variable_type, match(threshold, c("25%","35%","50%")), threshold) %>%
  mutate(across(c(pearson, spearman, reg_intercept, reg_slope, reg_r2), as.numeric))

kable(
  tab_s5_print,
  digits = 4,
  align = c("c","c","c","c"),
  col.names = c("Threshold", "Outcome", "N pairs", "Pearson r", "Spearman \u03c1",
                "Intercept", "Slope", "R\u00b2"),
  caption = "Robustness to non-votes: correlations and linear fits of filtered estimates on the full baseline (three-anchor dynIRT).",
   escape = FALSE
)
```

## SM3. Robustness to anchor choice

The main text relies on a multi-anchor solution to fix the orientation of the ideological scale. In Table S4 we document (i) which legislators were used as anchors, (ii) their availability and voting participation, and (iii) the extent to which alternative single-anchor solutions yield nearly identical ideal-point estimates (up to sign reversals). Because anchor choice can flip the orientation of the recovered scale, we align the sign of each single-anchor estimate to the baseline multi-anchor estimate before computing correlations and generating pairwise scatterplots.

```{r m3_anchors_setup, echo=FALSE, message=FALSE, warning=FALSE}

# ============================================================
# M2 — Robustness to anchor choice (single-anchor vs baseline)
# - Reads: candidates metadata, single-anchor estimates, baseline multi-anchor
# - Produces: Table S2 (anchor metadata), Table S3 (correlations), plots/heatmap elsewhere
# ============================================================

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(purrr)
  library(ggplot2)
  library(knitr)
  library(here)
})

aqui <- here()

# --- Paths ----------------------------------------------------
path_candidates <- paste0(aqui, "/pape/polarizacion/data/anchors_candidates_named.csv")
path_multi      <- paste0(aqui, "/pape/polarizacion/data/ideologia_diputados_largo_emIRT_multi_anclas.csv")

path_baseline_1 <- paste0(aqui, "/pape/polarizacion/data/base_completaideologia_diputados_largo_emIRT_con_anclas.csv")
path_baseline_2 <- paste0(aqui, "/pape/polarizacion/data/base_completa/ideologia_diputados_largo_emIRT_con_anclas.csv")
path_baseline   <- if (file.exists(path_baseline_1)) path_baseline_1 else path_baseline_2

cand  <- read_csv(path_candidates, show_col_types = FALSE)
multi <- read_csv(path_multi,      show_col_types = FALSE)
base3 <- read_csv(path_baseline,   show_col_types = FALSE)

# --- Helper: robust column detection ------------------------------------------
pick_first <- function(df, pattern) {
  nm  <- names(df)
  hit <- nm[str_detect(tolower(nm), pattern)]
  if (length(hit) == 0) NA_character_ else hit[1]
}

# Candidate metadata columns
col_id_cand   <- pick_first(cand, "^diputadoid$|diputado_id|id_diputado|^id$")
col_name_cand <- pick_first(cand, "nombre|name")
col_nfiles    <- pick_first(cand, "n_files_present|files_present|nfiles")
col_voterate  <- pick_first(cand, "vote_rate_overall|voterate|vote_rate")

if (is.na(col_id_cand) || is.na(col_nfiles) || is.na(col_voterate)) {
  stop("Could not identify DiputadoId / n_files_present / vote_rate_overall columns in anchors_candidates_named.csv.")
}

# Single-anchor panel identifiers
col_nombre_multi <- pick_first(multi, "^nombreid$|nombre_id|nombre")
col_period_multi <- pick_first(multi, "^periodo$|periodo_|period|term|session")
if (is.na(col_nombre_multi) || is.na(col_period_multi)) {
  stop("Could not identify NombreId / Periodo columns in ideologia_diputados_largo_emIRT_multi_anclas.csv.")
}

# Baseline identifiers and ideology column
col_nombre_base <- pick_first(base3, "^nombreid$|nombre_id|nombre")
col_period_base <- pick_first(base3, "^periodo$|periodo_|period|term|session")
col_ideol_base  <- pick_first(base3, "^ideologia$|^ideology$|ideal")
if (is.na(col_nombre_base) || is.na(col_period_base) || is.na(col_ideol_base)) {
  stop("Could not identify NombreId / Periodo / Ideologia columns in baseline file.")
}

# --- Detect single-anchor estimate columns -----------------------------------
ide_cols <- names(multi)[str_detect(names(multi), "^Ideologia_\\d+$")]
if (length(ide_cols) == 0) {
  stop("No columns of the form Ideologia_CCC were found in ideologia_diputados_largo_emIRT_multi_anclas.csv.")
}
anchor_ids_used <- as.integer(str_extract(ide_cols, "\\d+$"))

# Anchors highlighted in the main text
RIGHT_ID  <- 917L
CENTER_ID <- 879L
LEFT_ID   <- 837L

# --- Table S2: anchor metadata (+ force include the triad) --------------------
anchor_ids_tab <- sort(unique(c(anchor_ids_used, RIGHT_ID, CENTER_ID, LEFT_ID)))

anchor_type_map <- c(
  "917" = "right-wing",
  "879" = "center-moderate",
  "837" = "left-wing",
  "897" = "left-wing",
  "855" = "left-wing",
  "865" = "right-wing",
  "811" = "right-wing",
  "850" = "right-wing",
  "908" = "right-wing"
)

tab_anchor <- tibble(DiputadoId = anchor_ids_tab) %>%
  left_join(
    cand %>%
      transmute(
        DiputadoId        = as.integer(.data[[col_id_cand]]),
        anchor_name       = if (!is.na(col_name_cand)) as.character(.data[[col_name_cand]]) else NA_character_,
        n_files_present   = as.integer(.data[[col_nfiles]]),
        vote_rate_overall = as.numeric(.data[[col_voterate]])
      ),
    by = "DiputadoId"
  ) %>%
  mutate(
    anchor_type = dplyr::recode(as.character(DiputadoId), !!!anchor_type_map, .default = "other"),
    vote_rate_pct = if_else(!is.na(vote_rate_overall) & vote_rate_overall <= 1,
                            100 * vote_rate_overall, vote_rate_overall)
  ) %>%
  arrange(
    factor(anchor_type, levels = c("right-wing","center-moderate","left-wing","other")),
    desc(n_files_present),
    desc(vote_rate_pct)
  ) %>%
  select(DiputadoId, anchor_name, anchor_type, n_files_present, vote_rate_pct)

kable(
  tab_anchor,
  digits = 1,
  col.names = c("DiputadoId", "Name", "Anchor type", "Periods present (n)", "Vote rate overall (%)"),
  align   = c("c","l","c","c","c"),
  caption = "Anchor candidates used in single-anchor estimations (and baseline triad): availability across periods and overall voting participation."
)



```

Although the estimates are not identical, their correlations are all close to 1, as shown in Figure S1 below. Please pay attention to the range of values in the legend.

```{r fig_s1_heatmap, fig.width=8.5, fig.height=7, fig.cap="Correlation heatmap (sign-aligned). Color scale restricted to [0.90, 1.00].", fig.pos='H'}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(scales)
})

# --- Baseline and single-anchor estimates (LONG) ------------------------------
# Baseline multi-anchor estimate
base_long <- base3 %>%
  transmute(
    NombreId  = as.character(.data[[col_nombre_base]]),
    Periodo   = as.character(.data[[col_period_base]]),
    spec      = "Multi-anchor (baseline)",
    estimate  = as.numeric(.data[[col_ideol_base]]),
    anchor_id = NA_integer_
  )

# Single-anchor estimates
# FIX: select the Ideologia_CCC columns directly (do not splice character vectors)
multi_long <- multi %>%
  select(
    NombreId = all_of(col_nombre_multi),
    Periodo  = all_of(col_period_multi),
    all_of(ide_cols)
  ) %>%
  pivot_longer(
    cols = all_of(ide_cols),
    names_to = "spec_raw",
    values_to = "estimate"
  ) %>%
  mutate(
    estimate  = as.numeric(estimate),
    anchor_id = as.integer(str_extract(spec_raw, "\\d+$")),
    spec      = paste0("Anchor ", anchor_id)
  ) %>%
  select(NombreId, Periodo, spec, estimate, anchor_id)

# Combine (ensure both objects exist in-scope)
stopifnot(exists("base_long"), exists("multi_long"))

est_long <- bind_rows(base_long, multi_long) %>%
  arrange(NombreId, Periodo, spec, desc(!is.na(estimate))) %>%
  distinct(NombreId, Periodo, spec, .keep_all = TRUE)

# --- Sign alignment to the baseline ------------------------------------------
base_wide <- est_long %>%
  filter(spec == "Multi-anchor (baseline)") %>%
  select(NombreId, Periodo, base_est = estimate)

sign_map <- est_long %>%
  filter(spec != "Multi-anchor (baseline)") %>%
  inner_join(base_wide, by = c("NombreId","Periodo")) %>%
  filter(!is.na(estimate) & !is.na(base_est)) %>%
  group_by(spec) %>%
  summarise(
    n_pairs = n(),
    r_raw   = if (n() >= 2) cor(estimate, base_est) else NA_real_,
    sign_mult = if_else(!is.na(r_raw) & r_raw < 0, -1, 1),
    .groups = "drop"
  )

est_long_aligned <- est_long %>%
  left_join(sign_map %>% select(spec, sign_mult), by = "spec") %>%
  mutate(
    sign_mult = if_else(is.na(sign_mult), 1, sign_mult),
    estimate_aligned = estimate * sign_mult
  )

# NOTE: pair_data + corr tables + heatmap should use est_long_aligned downstream.

# Wide: one row per (NombreId, Periodo), columns = specs
wide_specs <- est_long_aligned %>%
  select(NombreId, Periodo, spec, estimate_aligned) %>%
  pivot_wider(names_from = spec, values_from = estimate_aligned)

spec_cols <- setdiff(names(wide_specs), c("NombreId", "Periodo"))

# Correlation matrix across specs
corr_mat <- cor(as.matrix(wide_specs[, spec_cols]), use = "pairwise.complete.obs")

# Long for ggplot
corr_long <- as.data.frame(as.table(corr_mat)) %>%
  transmute(
    spec_x = as.character(Var1),
    spec_y = as.character(Var2),
    r      = as.numeric(Freq)
  )

# Order + short labels (Multi, then anchor IDs)
spec_levels <- c("Multi-anchor (baseline)",
                 paste0("Anchor ", sort(as.integer(str_extract(spec_cols[str_detect(spec_cols, "^Anchor\\s+\\d+$")], "\\d+")))))
spec_levels <- spec_levels[spec_levels %in% spec_cols]

label_map <- setNames(
  c("Multi", str_extract(spec_levels[-1], "\\d+")),
  spec_levels
)

corr_long <- corr_long %>%
  mutate(
    spec_x = factor(spec_x, levels = spec_levels),
    spec_y = factor(spec_y, levels = spec_levels)
  )

p_heat <- ggplot(corr_long, aes(x = spec_x, y = spec_y, fill = r)) +
  geom_tile() +
  coord_fixed() +
  scale_x_discrete(labels = label_map) +
  scale_y_discrete(labels = label_map) +
  scale_fill_gradient(
    limits = c(0.99999, 1.00),
    oob    = scales::squish,
    breaks = c(0.99999, 1.00),
    labels = function(x) gsub("\\.", ",", formatC(x, format = "f", digits = 5))
  ) +
  labs(x = NULL, y = NULL, fill = "r") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid  = element_blank()
  )

print(p_heat)

```

## SM4. Estimation uncertainty and uncertainty propagation

To propagate legislator–period ideal-point uncertainty into period-level polarization, we use the 95% intervals reported by dynIRT. Treating $[x_{it}^{low}, x_{it}^{high}]$ as an approximate 95% interval around the point estimate $\hat{x}{it}$, we recover $\hat{\sigma}{it} \approx (x_{it}^{high}-x_{it}^{low})/(2\cdot 1.96)$ under a normal approximation. We then simulate $x_{it}^{(b)}\sim \mathcal{N}(\hat{x}{it},\hat{\sigma}{it}^2)$ for $b=1,\dots,B$, recompute period-level polarization for each draw, and report uncertainty bands as the 2.5th and 97.5th percentiles of the draw-level polarization distribution.

Table S5 (median and IQR of interval widths by period) shows that ideal-point uncertainty is highly homogeneous within legislatures: variation in interval widths across legislators within a period is negligible, so any differences in precision arise primarily across periods rather than from a small subset of unusually imprecise legislators. Consistent with this pattern, the propagated uncertainty bands for polarization are narrow and do not affect the ordering of periods or the substantive interpretation of the polarization trajectory.

```{r m4_1, message=FALSE, warning=FALSE, echo=FALSE}
# ============================================================
# M4 — Uncertainty & propagation to polarization measures
# Inputs: a legislator-period file that includes point estimates + *_low/*_high bands.
# Output objects:
#   - tab_uncert_pol: period-level polarization with propagated uncertainty intervals
#   - fig_uncert_ip:  distribution of ideal-point interval widths by period
#   - fig_uncert_pol: polarization trajectories with uncertainty ribbons
# ============================================================

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(purrr)
  library(ggplot2)
  library(knitr)
  library(here)
})

aqui <- here()

# --- 1) Prefer the *output* file produced by your non-vote sensitivity script (it has IC_low/IC_high)
candidate_paths <- c(file.path(aqui, "pape/polarizacion/data/ideologia_diputados_largo_emIRT_con_anclas.csv"))

path_ip <- candidate_paths[file.exists(candidate_paths)][1]
if (is.na(path_ip)) {
  stop("M4: No candidate ideal-point file found. Checked:\n", paste(candidate_paths, collapse = "\n"))
}

ip <- read_csv(path_ip, show_col_types = FALSE)

pick_first <- function(df, pattern) {
  nm <- names(df)
  hit <- nm[str_detect(tolower(nm), pattern)]
  if (length(hit) == 0) NA_character_ else hit[1]
}

# --- 2) Detect columns (baseline dynIRT outputs are usually: DiputadoId, NombreId, Periodo, IC_low, Ideologia, IC_high)
col_id     <- pick_first(ip, "^diputadoid$|^nombreid$|nombre_id|diputado_id|^id$")
col_period <- pick_first(ip, "^periodo$|period_num|period(_id)?|legisl|term|session")
col_mean   <- pick_first(ip, "^ideologia$|^ideology$|^ideal$|^estimate$")
col_low    <- pick_first(ip, "(_low$|^ic_low$|^ci_low$|lower$|lwr$)")
col_high   <- pick_first(ip, "(_high$|^ic_high$|^ci_high$|upper$|upr$)")

if (any(is.na(c(col_id, col_period, col_mean, col_low, col_high)))) {
  stop(
    "M4: Could not detect required columns in: ", path_ip, "\n",
    "Detected: id=", col_id, " period=", col_period, " mean=", col_mean,
    " low=", col_low, " high=", col_high, "\n\n",
    "Columns available:\n", paste(names(ip), collapse = ", ")
  )
}

# --- 2) Robust period parsing (handles '1' and 'Periodo_1' etc.)
parse_period <- function(x) {
  x <- as.character(x)
  as.integer(str_extract(x, "\\d+"))
}


# --- 2) Build clean legislator-period panel with ordered bounds ---------------
ip2 <- ip %>%
  transmute(
    id = as.character(.data[[col_id]]),

    # FIX: robust period parsing (handles "Periodo_1", "1", "2002-06", etc.)
    period_raw = as.character(.data[[col_period]]),
    period = suppressWarnings(as.integer(period_raw)),
    period = dplyr::coalesce(period, suppressWarnings(as.integer(stringr::str_extract(period_raw, "\\d+")))),

    # Robust numeric parsing (works even if some columns are characters)
    x_hat    = suppressWarnings(readr::parse_number(as.character(.data[[col_mean]]))),
    low_raw  = suppressWarnings(readr::parse_number(as.character(.data[[col_low]]))),
    high_raw = suppressWarnings(readr::parse_number(as.character(.data[[col_high]])))
  ) %>%
  filter(!is.na(period), !is.na(x_hat), !is.na(low_raw), !is.na(high_raw)) %>%
  mutate(
    # enforce low <= high even if inverted
    x_low  = pmin(low_raw,  high_raw),
    x_high = pmax(low_raw,  high_raw)
  ) %>%
  select(id, period, x_hat, x_low, x_high)


# --- 3) Recover SD from (approx.) 95% interval --------------------------------
z975 <- qnorm(0.975)
ip2 <- ip2 %>%
  mutate(
    ci_width = x_high - x_low,
    x_sd     = ci_width / (2 * z975)
  ) %>%
  filter(is.finite(x_sd), x_sd >= 0)

# Period labels (edit if needed)
period_labels <- c("1"="2002-06","2"="2006-10","3"="2010-14","4"="2014-18","5"="2018-22","6"="2022-26")
ip2 <- ip2 %>% mutate(period_lab = recode(as.character(period), !!!period_labels))

# --- 4) Propagation via Monte Carlo draws -------------------------------------
set.seed(123)
B_draws <- 2000L  # number of Monte Carlo draws (integer)

calc_measures_one_period <- function(df_p, B_draws) {

  # Defensive typing (prevents "non-numeric" errors)
  B_draws <- as.integer(B_draws)
  n <- as.integer(nrow(df_p))

  if (is.na(B_draws) || B_draws < 2) stop("B_draws must be a positive integer >= 2.")
  if (is.na(n) || n < 1) stop("Empty period: no rows in df_p.")

  xhat <- as.numeric(df_p$x_hat)
  xsd  <- as.numeric(df_p$x_sd)

  # Draw matrix: rows = legislators, cols = draws
  X <- matrix(
    rnorm(n * B_draws,
          mean = rep(xhat, times = B_draws),
          sd   = rep(xsd,  times = B_draws)),
    nrow = n, ncol = B_draws, byrow = FALSE
  )

  abs_draw      <- colMeans(abs(X), na.rm = TRUE)
  sq_draw       <- colMeans(X^2, na.rm = TRUE)
  med_draw      <- apply(X, 2, median, na.rm = TRUE)
  abs_med_draw  <- colMeans(abs(X - matrix(med_draw, nrow = n, ncol = B_draws, byrow = TRUE)), na.rm = TRUE)

  tibble(
    draw = 1:B_draws,
    abs = abs_draw,
    sq = sq_draw
  )
}

draws_by_period <- ip2 %>%
  group_by(period, period_lab) %>%
  group_modify(~ calc_measures_one_period(.x, B_draws = B_draws)) %>%
  ungroup()

# --- 5) tab_uncert_pol: summarize draw distributions per period ---------------
summ_one <- function(x) {
  if (all(is.na(x))) return(c(est = NA_real_, low = NA_real_, high = NA_real_))
  est <- mean(x, na.rm = TRUE)
  qs  <- as.numeric(quantile(x, probs = c(0.025, 0.975), na.rm = TRUE, names = FALSE))
  c(est = est, low = qs[1], high = qs[2])
}

tab_uncert_pol <- draws_by_period %>%
  group_by(period, period_lab) %>%
  summarise(across(c(abs, sq), ~ list(summ_one(.x))), .groups="drop") %>%
  pivot_longer(cols = c(abs, sq), names_to = "measure", values_to = "tmp") %>%
  mutate(
    est  = map_dbl(tmp, ~ .x[["est"]]),
    low  = map_dbl(tmp, ~ .x[["low"]]),
    high = map_dbl(tmp, ~ .x[["high"]]),
    measure = recode(measure,
                     abs = "abs",
                     sq = "sq")
  ) %>%
  select(period, period_lab, measure, est, low, high) %>%
  arrange(match(period, 1:6), factor(measure, levels=c("abs","sq")))

# probability that the linear trend is positive (one scalar per measure)
trend_prob <- draws_by_period %>%
  pivot_longer(cols = c(abs, sq),
               names_to = "measure", values_to = "pol") %>%
  group_by(measure, draw) %>%
  summarise(
    slope = {
      t <- as.numeric(period)                     # time within this group
      y <- as.numeric(pol)

      cc <- complete.cases(t, y)
      if (sum(cc) < 2) {
        NA_real_
      } else {
        # slope = cov(t,y)/var(t) computed on matched, non-missing pairs
        cov(t[cc], y[cc]) / var(t[cc])
      }
    },
    .groups = "drop"
  ) %>%
  group_by(measure) %>%
  summarise(p_slope_pos = mean(slope > 0, na.rm = TRUE), .groups = "drop")

# trend_prob
```

```{r m4_2, message=FALSE, warning=FALSE, echo=FALSE}
p_ip <- ip2 %>%
  ggplot(aes(x = period_lab, y = ci_width)) +
  geom_boxplot(outlier.size = 0.6) +
  labs(x = NULL, y = "Ideal-point interval width") +
  theme_minimal()

#print(p_ip)

# ip2c <- ip2 %>%
#   group_by(period_lab) %>%
#   mutate(ci_center = (ci_width - mean(ci_width)) * 1e6) %>%  # en “micro-units”
#   ungroup()
#
# ggplot(ip2c, aes(period_lab, ci_center)) +
#   geom_boxplot(outlier.size = 0.6) +
#   labs(x = NULL, y = "Width deviation from period mean (×10^6)") +
#   theme_minimal()

ip2 %>%
  group_by(period_lab) %>%
  summarise(
    median = median(ci_width),
    p25    = quantile(ci_width, 0.25),
    p75    = quantile(ci_width, 0.75),
    min    = min(ci_width),
    max    = max(ci_width),
    .groups = "drop"
  ) %>%
  arrange(period_lab) %>%
  knitr::kable(
    format  = "latex",
    digits  = 3,
    booktabs = TRUE,
    caption = "Legislator-level ideal-point uncertainty by period (CI width summary)."
  )



```

The uncertainty intervals in Table S6 quantify how ideal-point uncertainty propagates into period-level polarization summaries. For each period, we take the estimated legislator ideal points $\hat{x}_{it}$ and recover an approximate standard deviation from the reported 95% interval $[x^{\text{low}}_{it}, x^{\text{high}}_{it}]$. We then draw Monte Carlo realizations $x^{(b)}_{it} \sim \mathcal{N}(\hat{x}_{it}, \hat{\sigma}_{it}^2)$ and recompute two polarization measures in each draw: mean absolute ideal point (*abs*) and mean squared ideal point (*sq*). The table reports the Monte Carlo mean (“Estimate”) and the central 95% interval (2.5%–97.5%) of each measure’s sampling distribution induced solely by ideal-point uncertainty.

Two patterns are evident. First, the propagated intervals are very narrow in every period and for every measure (typically a few thousandths wide), indicating that the main time-series movements in polarization are orders of magnitude larger than the uncertainty introduced by estimation error in individual ideal points. Second, the time-profile of each measure remains unchanged once uncertainty is propagated: polarization rises from early periods to 2018–22 and remains high in 2022–26, regardless of whether polarization is summarized by absolute positions, deviations from central tendency, or squared positions.

```{r m4_3, message=FALSE, warning=FALSE, echo=FALSE}
# Table: propagated uncertainty intervals for polarization by period
# Formatting: 4 decimals and centered columns

fmt4 <- function(x) gsub("\\.", ",", formatC(x, format="f", digits=4))

tab_print <- tab_uncert_pol %>%
  mutate(
    est  = fmt4(est),
    low  = fmt4(low),
    high = fmt4(high),
    measure = recode(measure,
                     abs = "abs",
                     sq = "sq")
  ) %>%
  select(`Period` = period_lab, `Measure` = measure, `Estimate` = est, `Low (2.5%)` = low, `High (97.5%)` = high)

kable(
  tab_print,
  align = rep("c", ncol(tab_print)),
  caption = "Uncertainty-propagated polarization intervals by period (Monte Carlo propagation from ideal-point uncertainty)."
)
```

## SM5. Additional Implementations and Diagnostics (Model Stability Across Designs)

```{r m5_1, message=FALSE, warning=FALSE, echo=FALSE}
# ============================================================
# M5_1 — Inventory + standardized coefficient extraction
# Goal: locate all "polarizacion_dynamic_*_coeficientes.csv" under:
#   root/{base_completa, subsamples_bill_origin, subsamples_vote_step, submuestras_balanceadas/resumen}
# and build a clean long table of key coefficients for:
#   period trend (period_num), elite replacement (group_tNuevo_en_t),
#   post-reform (post_reform), within-legislator change since entry (tau)
# ============================================================

suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(stringr)
  library(purrr)
  library(tidyr)
  library(here)
})

aqui <- here()

# --- 0) Root (support both spellings) ----------------------------------------
root_candidates <- c(
  file.path(aqui, "pape", "polarizacion", "data"),
  file.path(aqui, "pape", "polarization", "data")
)
root <- root_candidates[dir.exists(root_candidates)][1]
if (is.na(root)) {
  stop("M5_1: Could not find root data directory. Tried: ",
       paste(root_candidates, collapse = " | "))
}

# --- 1) Where to search -------------------------------------------------------
dir_map <- tibble::tribble(
  ~design,       ~base,
  "full",        file.path(root, "base_completa"),
  "bill_origin", file.path(root, "subsamples_bill_origin"),
  "vote_step",   file.path(root, "subsamples_vote_step"),
  "balanced",    file.path(root, "submuestras_balanceadas", "resumen")
) %>%
  mutate(exists = dir.exists(base))

if (!any(dir_map$exists)) {
  stop("M5_1: None of the expected design directories exist under root = ", root)
}

# --- 2) Helpers ----------------------------------------------------------------
find_pol_dirs <- function(base_dir) {
  if (!dir.exists(base_dir)) return(character(0))
  dd <- list.dirs(base_dir, recursive = TRUE, full.names = TRUE)
  dd[tolower(basename(dd)) == "polarizacion"]
}

detect_pol_def <- function(fname) {
  # vector-safe
  defs <- c("sq_ideology", "abs") # longer first
  fn <- as.character(fname)

  out <- rep(NA_character_, length(fn))
  for (d in defs) {
    pat <- paste0("(^|_)", d, "(_|\\.|$)")
    hit <- stringr::str_detect(fn, pat)
    out[is.na(out) & hit] <- d
  }
  out
}

std_model_name <- function(x, fallback = NA_character_) {
  z <- tolower(as.character(x))
  out <- dplyr::case_when(
    str_detect(z, "model\\s*1|\\bm1\\b") ~ "M1",
    str_detect(z, "model\\s*2|\\bm2\\b") ~ "M2",
    str_detect(z, "model\\s*3|\\bm3\\b") ~ "M3",
    str_detect(z, "model\\s*4|\\bm4\\b") ~ "M4",
    str_detect(z, "model\\s*5|\\bm5\\b") ~ "M5",
    str_detect(z, "model\\s*6|\\bm6\\b") ~ "M6",
    TRUE ~ NA_character_
  )
  ifelse(is.na(out), fallback, out)
}

pick_first <- function(nms, pattern) {
  # nms = names(df)
  hit <- nms[str_detect(tolower(nms), pattern)]
  if (length(hit) == 0) NA_character_ else hit[1]
}

read_coefs_standardized <- function(path, model_hint = NA_character_) {
  df <- tryCatch(
    read_csv(path, show_col_types = FALSE, progress = FALSE),
    error = function(e) NULL
  )
  if (is.null(df) || nrow(df) == 0) return(tibble())

  nms <- names(df)

  # detect columns (robust)
  col_term  <- pick_first(nms, "^(term|variable|predictor|parameter|coef)$")
  col_est   <- pick_first(nms, "^(estimate|est|beta|b)$")
  col_se    <- pick_first(nms, "std\\.error|std_error|stderr|se")
  col_p     <- pick_first(nms, "^p\\.value$|p_value|pvalue")
  col_star  <- pick_first(nms, "stars")
  col_model <- pick_first(nms, "^model$|^modelo$|spec|specification")

  # if we cannot find term+estimate, we can't use this file
  if (is.na(col_term) || is.na(col_est)) return(tibble())

  out <- df %>%
    transmute(
      model_raw  = if (!is.na(col_model)) as.character(.data[[col_model]]) else NA_character_,
      term       = as.character(.data[[col_term]]),
      estimate   = suppressWarnings(as.numeric(.data[[col_est]])),
      std_error  = if (!is.na(col_se)) suppressWarnings(as.numeric(.data[[col_se]])) else NA_real_,
      p_value    = if (!is.na(col_p))  suppressWarnings(as.numeric(.data[[col_p]]))  else NA_real_,
      stars      = if (!is.na(col_star)) as.character(.data[[col_star]]) else NA_character_
    ) %>%
    mutate(
      term  = str_replace_all(term, "`", ""),
      model = std_model_name(model_raw, fallback = model_hint)
    ) %>%
    select(model, term, estimate, std_error, p_value, stars)

  out
}

# --- 3) Collect all coefficient files -----------------------------------------
pol_dirs <- dir_map %>%
  filter(exists) %>%
  mutate(pol_dirs = map(base, find_pol_dirs)) %>%
  select(design, base, pol_dirs) %>%
  unnest(pol_dirs)

if (nrow(pol_dirs) == 0) {
  stop("M5_1: No directories named 'polarizacion' were found under: ",
       paste(dir_map$base[dir_map$exists], collapse = " | "))
}

coef_files <- pol_dirs %>%
  mutate(files = map(pol_dirs, ~ list.files(
    .x,
    pattern = "^polarizacion_dynamic_.*_coeficientes\\.csv$",
    full.names = TRUE
  ))) %>%
  select(design, base, pol_dirs, files) %>%
  unnest(files) %>%
  rename(coef_path = files)

if (nrow(coef_files) == 0) {
  stop("M5_1: Found 'polarizacion' dirs, but no files matching 'polarizacion_dynamic_*_coeficientes.csv'.")
}

# --- 4) Parse metadata from paths ---------------------------------------------
coef_meta <- coef_files %>%
  mutate(
    file      = basename(coef_path),
    normalized = str_detect(coef_path, "(/|\\\\)Normalized(/|\\\\)"),
    # sample label = directory right above /polarizacion
    sample_raw = basename(dirname(pol_dirs)),
    sample = case_when(
      design == "full"        ~ "Full sample (integrated)",
      design == "balanced"    ~ "Balanced subsample (mean 100 runs)",
      design == "bill_origin" ~ paste0("Bill origin: ", sample_raw),
      design == "vote_step"   ~ paste0("Vote step: ", sample_raw),
      TRUE ~ sample_raw
    ),
    pol_def = detect_pol_def(file),
    model_hint = case_when(
      str_detect(tolower(file), "model5") ~ "M5",
      str_detect(tolower(file), "model6") ~ "M6",
      TRUE ~ NA_character_
    )
  )

# --- 5) Read and standardize all coefficients ---------------------------------
coefs_long <- purrr::pmap_dfr(
  coef_meta %>% select(coef_path, design, sample, pol_def, normalized, model_hint),
  function(coef_path, design, sample, pol_def, normalized, model_hint) {
    read_coefs_standardized(coef_path, model_hint = model_hint) %>%
      mutate(
        coef_path = coef_path,
        design    = design,
        sample    = sample,
        pol_def   = pol_def,
        normalized = normalized
      )
  }
)

if (nrow(coefs_long) == 0) {
  stop("M5_1: Coefficient files were found, but none had recognizable (term, estimate) columns.")
}

# --- 6) Keep only the terms we promised to summarize in M5 --------------------
# Required names (as you specified)
key_regex <- paste0(
  "(",
  "^(period_num)$|",
  "^(tau)$|",
  "^group_tNuevo_en_t|",
  "^post_reform|",
  # common interactions (either order)
  "period_num:group_tNuevo_en_t|group_tNuevo_en_t:period_num|",
  "post_reform:period_num|period_num:post_reform|",
  "post_reform:group_tNuevo_en_t|group_tNuevo_en_t:post_reform|",
  "post_reform:tau|tau:post_reform",
  ")"
)

coefs_key <- coefs_long %>%
  filter(!is.na(term)) %>%
  filter(str_detect(term, key_regex)) %>%
  mutate(
    term_label = case_when(
      term == "period_num" ~ "Legislative period trend",
      term == "tau" ~ "Within-legislator change since entry (tau)",
      str_detect(term, "^group_tNuevo_en_t") ~ "Elite replacement (group_tNuevo_en_t)",
      str_detect(term, "^post_reform") ~ "Post-reform term (post_reform)",
      str_detect(term, "period_num:group_tNuevo_en_t|group_tNuevo_en_t:period_num") ~ "Replacement × period trend",
      str_detect(term, "post_reform:period_num|period_num:post_reform") ~ "Post-reform × period trend",
      str_detect(term, "post_reform:group_tNuevo_en_t|group_tNuevo_en_t:post_reform") ~ "Post-reform × replacement",
      str_detect(term, "post_reform:tau|tau:post_reform") ~ "Post-reform × tau",
      TRUE ~ term
    )
  )

# If this fails, it means naming differs in your CSV outputs; we want to know now.
if (nrow(coefs_key) == 0) {
  stop(
    "M5_1: coefs_long loaded, but none of the key terms were found.\n",
    "Searched regex = ", key_regex, "\n",
    "Example terms present (first 30): ",
    paste(head(unique(coefs_long$term), 30), collapse = " | ")
  )
}

# --- 7) Small inventory (used later by M5 chunks) -----------------------------
m5_inventory <- coefs_long %>%
  distinct(design, sample, pol_def, normalized, model) %>%
  arrange(design, sample, pol_def, normalized, model)

# Objects exported for subsequent chunks:
#   - coefs_long     (all standardized coefficients)
#   - coefs_key      (only the key terms for M5 synthesis)
#   - m5_inventory   (what was found)
```

To assess whether the upward trend documented in the main text depends on the specific polariztion metric, we plot it again in Figure S2 together with an alternative operationalization from the same ideal-point estimates and normalized to the first period (2002–06 = 1). The left panel reports mean absolute ideological position, $\mathbb{E}(|x|)$ (*abs*), which tracks average dispersion from the reference point in a linear way; the right panel reports mean squared ideological position, $\mathbb{E}(x^2)$ (*sq*), which places disproportionate weight on more extreme legislators. Both measures reproduce the same qualitative dynamics as the baseline series in the main text: polarization rises steadily from the early periods through 2018–22 and remains elevated in 2022–26. The stronger increase under the quadratic measure indicates that the overall rise is not driven solely by moderate dispersion, but also by a growing contribution of legislators located farther in the tails of the ideological distribution.

```{r m5_2, message=FALSE, warning=FALSE, echo=FALSE, fig.cap="Trends in polarization metrics", fig.pos='H'}

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)
  library(here)
})

aqui <- here()

# --- Helper: robust column detection ------------------------------------------
pick_first <- function(df, pattern) {
  nm  <- names(df)
  hit <- nm[str_detect(tolower(nm), pattern)]
  if (length(hit) == 0) NA_character_ else hit[1]
}

# --- 1) Locate baseline ideal-points file (try common locations) --------------
candidates <- c(
  file.path(aqui, "pape/polarizacion/data/base_completa/ideologia_diputados_largo_emIRT_con_anclas.csv"),
  file.path(aqui, "pape/polarizacion/data/ideologia_diputados_largo_emIRT_con_anclas.csv"),
  file.path(aqui, "pape/polarizacion/data/ideologia_diputados_largo_emIRT_con_anclas_normalized.csv"),
  file.path(aqui, "pape/polarizacion/data/base_completa/ideologia_diputados_largo_emIRT_con_anclas_normalized.csv")
)
path_ip <- candidates[file.exists(candidates)][1]

if (is.na(path_ip)) {
  cat("**Figure S?:** Baseline ideal-points file not found. Checked:\n")
  cat(paste0("- ", candidates, collapse = "\n"))
} else {

  ip <- read_csv(path_ip, show_col_types = FALSE)

  col_period <- pick_first(ip, "^periodo$|^period$|period_num|period(_id)?|legisl|term|session")
  col_x      <- pick_first(ip, "^ideologia$|^ideology$|ideal(?!.*low)(?!.*high)")

  if (any(is.na(c(col_period, col_x)))) {
    cat("**Figure S?:** Could not detect required columns in baseline file.\n")
    cat("- period column:", col_period, "\n")
    cat("- ideology column:", col_x, "\n")
  } else {

    # --- 2) Build clean panel: (period_num, x) --------------------------------
    ip2 <- ip %>%
      transmute(
        period_raw = as.character(.data[[col_period]]),
        x = suppressWarnings(as.numeric(.data[[col_x]]))
      ) %>%
      mutate(
        period_num = suppressWarnings(as.integer(period_raw)),
        period_num = ifelse(
          is.na(period_num) & str_detect(period_raw, "\\d+"),
          suppressWarnings(as.integer(str_extract(period_raw, "\\d+"))),
          period_num
        )
      ) %>%
      filter(!is.na(period_num), is.finite(x))

    if (nrow(ip2) == 0) {
      cat("**Figure S?:** No usable rows after cleaning (period/x).\n")
    } else {

      # Period labels (edit if your mapping differs)
      period_labels <- c("1"="2002-06","2"="2006-10","3"="2010-14","4"="2014-18","5"="2018-22","6"="2022-26")

      # --- 3) Compute alternative polarization measures by period --------------
      pol_by_period <- ip2 %>%
        group_by(period_num) %>%
        summarise(
          n = n(),
          mu = mean(x, na.rm = TRUE),
          med = median(x, na.rm = TRUE),
          abs = mean(abs(x), na.rm = TRUE),
          sq = mean(x^2, na.rm = TRUE),
          .groups = "drop"
        ) %>%
        mutate(
          period_lab = dplyr::recode(as.character(period_num), !!!period_labels),
          period_lab = ifelse(is.na(period_lab), as.character(period_num), period_lab)
        ) %>%
        arrange(period_num)

      # --- 4) Index to first period to make magnitudes comparable --------------
      pol_long <- pol_by_period %>%
        select(period_num, period_lab, abs, sq) %>%
        pivot_longer(cols = c(abs, sq),
                     names_to = "measure", values_to = "value") %>%
        group_by(measure) %>%
        mutate(
          base0 = value[period_num == min(period_num, na.rm = TRUE)][1],
          index = value / base0
        ) %>%
        ungroup() %>%
        mutate(
          measure = factor(measure, levels = c("abs","sq"),
                           labels = c("Mean |x| (abs)",
                                      "Mean x² (sq)"))
        )

      # --- 5) Plot (compact + robust) -----------------------------------------
      p_alt <- ggplot(pol_long, aes(x = period_num, y = index, group = measure)) +
        geom_line(linewidth = 0.5) +
        geom_point(size = 1.0) +
        facet_wrap(~ measure, ncol = 2) +
        scale_x_continuous(
          breaks = pol_by_period$period_num,
          labels = pol_by_period$period_lab
        ) +
        labs(x = NULL, y = "Index (2002–06 = 1)") +
        theme_minimal() +
        theme(
          panel.grid.minor = element_blank(),
          axis.text.x = element_text(angle = 45, hjust = 1)
        )

      print(p_alt)
    }
  }
}

```

#### On the stability of key coefficients in the regressions

Table S7 summarizes *sign and statistical stability* of the key coefficients across designs and alternative dependent variables. For each term (Legislative period trend = `period_num`, Elite replacement = `group_t` equals 1 when the individual is new comer in period t, Post-reform = `post_reform`, and Within-legislator trend = Δ since entry), each DV definition (here: `abs` and `sq_ideology`), and each design family (A to D as described below), we count how many subsample-specific coefficient estimates were computed (`N`) and report the share whose 95% confidence interval excludes 0 versus the share whose 95% interval crosses 0 . Within each (term × DV × design) cell, these two shares sum to 100%.

We estimate the six models using the full roll-call dataset and a set of subsamples organized into four categories, labeled A through D. Group A corresponds to the full sample. Group B stratifies votes by bill origin, distinguishing between executive- and legislatively initiated bills. Group C covers procedural stages, including first, second, third, and mixed committee stages. Group D summarizes results from 100 draws of balanced samples constructed by holding constant the total number of votes across periods.

First, Legislative period trend and Within-legislator trend (both, alternative operationalizations of Elite adaptation) are the most stable results across all designs, *all* available subsample estimates exclude 0 for both `abs` and `sq`, indicating that the upward time pattern and within-legislator drift are not driven by any single subsampling strategy. Second, Elite replacement is also highly stable, but weakens modestly under the vote-step design (C), where a non-trivial minority of estimates cross 0 (e.g., \~25%), consistent with the idea that splitting by legislative stage reduces precision and/or introduces greater heterogeneity without overturning the overall pattern. Third, Post-reform is the least stable term: in several design–DV combinations it mostly (or entirely) crosses 0, and where it excludes 0 it does so only for a minority of subsamples. Substantively, this supports treating the reform-related coefficient as *more design-sensitive and less precisely identified* than the legislative period trend, replacement, and within-legislator trend.

```{r m5_3, message=FALSE, warning=FALSE, echo=FALSE}
# ============================================================
# M5_3 — Key regression coefficients across modules
# Goal:
#   1) Build coefs_all directly from discovered *_coeficientes.csv files
#   2) Produce a compact diagnostic table (NOT huge) and a figure
# Notes:
#   - Searches under the common root (with fallbacks)
#   - Handles different column schemas (term/estimate/se or cell parsing)
#   - Avoids rendering errors if some dirs/files are missing
# ============================================================

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(purrr)
  library(ggplot2)
  library(knitr)
  library(here)
})

aqui <- here()

# --- 0) Robust roots (your stated root + fallbacks) ---------------------------
roots <- c(
  file.path(aqui, "pape", "polarization", "data"),
  file.path(aqui, "pape", "polarizacion", "data"),
  file.path(aqui, "pape", "polarizacion", "output"),
  file.path(aqui, "pape", "polarization", "output")
)
roots <- roots[dir.exists(roots)]
stopifnot(length(roots) >= 1)

# --- 1) Helpers ---------------------------------------------------------------
pick_first <- function(df, pattern) {
  nm  <- names(df)
  hit <- nm[str_detect(tolower(nm), pattern)]
  if (length(hit) == 0) NA_character_ else hit[1]
}

norm_path <- function(x) gsub("\\\\", "/", x)

detect_family <- function(path) {
  p <- norm_path(path)
  dplyr::case_when(
    str_detect(p, "/subsamples_bill_origin/")      ~ "subsamples_bill_origin",
    str_detect(p, "/subsamples_vote_step/")        ~ "subsamples_vote_step",
    str_detect(p, "/submuestras_balanceadas/")     ~ "submuestras_balanceadas",
    str_detect(p, "/base_completa/")               ~ "base_completa",
    TRUE ~ "other"
  )
}

extract_sample <- function(path, family) {
  p <- norm_path(path)
  parts <- str_split(p, "/", simplify = TRUE)
  parts <- as.character(parts[parts != ""])
  i <- match(family, parts)
  if (is.na(i)) return(NA_character_)

  # Balanced samples: .../submuestras_balanceadas/resumen/<sample>/polarizacion/...
  if (family == "submuestras_balanceadas") {
    j <- match("resumen", parts)
    if (!is.na(j) && length(parts) >= j + 1) return(parts[j + 1])
  }

  # General: take next folder after the family key (often the subsample label)
  if (length(parts) >= i + 1) return(parts[i + 1])
  NA_character_
}

detect_pol_def <- function(path) {
  fname <- tolower(basename(path))
  defs <- c("sq_ideology", "abs")

  # IMPORTANT: do NOT pass a vector of patterns to str_detect()
  hits <- defs[vapply(defs, function(d) {
    str_detect(fname, paste0("(^|_)", d, "(_|\\.|$)"))
  }, logical(1))]

  if (length(hits) == 0) NA_character_ else hits[1]
}

detect_model_tag <- function(path) {
  fname <- tolower(basename(path))
  dplyr::case_when(
    str_detect(fname, "model5") ~ "model5",
    str_detect(fname, "model6") ~ "model6",
    TRUE ~ "main_models"
  )
}

fmt4 <- function(x) {
  ifelse(is.na(x), "", gsub("\\.", ",", formatC(x, format = "f", digits = 4)))
}

# Parse estimate/se from a "cell" string if needed: "0.123*** (0.045)"
parse_cell <- function(cell) {
  cell <- as.character(cell)
  est <- suppressWarnings(as.numeric(str_extract(cell, "^-?\\d+\\.?\\d*")))
  se  <- suppressWarnings(as.numeric(str_extract(cell, "(?<=\\()\\s*-?\\d+\\.?\\d*(?=\\))")))
  tibble(estimate = est, std_error = se)
}

safe_read <- purrr::possibly(
  function(f) read_csv(f, show_col_types = FALSE),
  otherwise = NULL
)

# --- 2) Find coefficient files ------------------------------------------------
coef_files <- unlist(lapply(roots, function(r) {
  list.files(r, recursive = TRUE, full.names = TRUE, pattern = "coeficientes\\.csv$", ignore.case = TRUE)
}))

coef_files <- coef_files %>%
  norm_path() %>%
  unique() %>%
  keep(~ str_detect(.x, "polarizacion_dynamic_"))

if (length(coef_files) == 0) {
  cat("**M5_3:** No coefficient files found under the roots. Check the root paths and naming.\n")
} else {

  # --- 3) Read + standardize into coefs_all ----------------------------------
  coefs_all <- map_dfr(coef_files, function(f) {
    df <- safe_read(f)
    if (is.null(df) || nrow(df) == 0) return(NULL)

    col_term <- pick_first(df, "^term$|^variable$|^param|^parameter$")
    col_est  <- pick_first(df, "^estimate$|^estim|^beta$|^coef$")
    col_se   <- pick_first(df, "std\\.error|std_error|^se$|stderr")
    col_mod  <- pick_first(df, "^model$|^modelo$|spec|specification")
    col_cell <- pick_first(df, "^cell$")

    out <- df %>%
      transmute(
        file      = f,
        family    = detect_family(f),
        sample    = extract_sample(f, detect_family(f)),
        normalized = str_detect(norm_path(f), "/normalized/"),
        pol_def   = detect_pol_def(f),
        model_tag = detect_model_tag(f),
        model     = if (!is.na(col_mod)) as.character(.data[[col_mod]]) else NA_character_,
        term      = if (!is.na(col_term)) as.character(.data[[col_term]]) else NA_character_,
        estimate  = if (!is.na(col_est)) suppressWarnings(as.numeric(.data[[col_est]])) else NA_real_,
        std_error = if (!is.na(col_se))  suppressWarnings(as.numeric(.data[[col_se]]))  else NA_real_,
        cell      = if (!is.na(col_cell)) as.character(.data[[col_cell]]) else NA_character_
      )

    # If estimate/se are missing but we have cell, parse them
    if (all(is.na(out$estimate)) && !all(is.na(out$cell))) {
      parsed <- parse_cell(out$cell)
      out$estimate  <- parsed$estimate
      out$std_error <- parsed$std_error
    }

    out %>%
      select(-cell) %>%
      filter(!is.na(term)) %>%
      mutate(
        family = if_else(is.na(family), "other", family),
        sample = if_else(is.na(sample), "NA", sample),
        pol_def = if_else(is.na(pol_def), "unknown", pol_def)
      )
  })

  stopifnot(exists("coefs_all"), nrow(coefs_all) > 0)

  # --- 4) Focus on the four key terms you specified ---------------------------
  key_terms <- c("period_num", "group_tNuevo_en_t", "post_reform", "tau")
  coefs_all <- coefs_all %>% filter(!normalized)

# estandariza etiqueta de modelo usando `model` si existe y si no, el filename
std_model <- function(model, file) {
  m <- tolower(ifelse(is.na(model), "", model))
  f <- tolower(basename(file))
  dplyr::case_when(
    stringr::str_detect(m, "\\bm4\\b|model\\s*4") ~ "M4",
    stringr::str_detect(m, "\\bm5\\b|model\\s*5") ~ "M5",
    stringr::str_detect(m, "\\bm6\\b|model\\s*6") ~ "M6",
    stringr::str_detect(f, "model5") ~ "M5",
    stringr::str_detect(f, "model6") ~ "M6",
    TRUE ~ NA_character_
  )
}

key_terms <- c("period_num", "group_tNuevo_en_t", "post_reform", "tau")

coefs_key <- coefs_all %>%
  mutate(model_std = std_model(model, file)) %>%
  filter(term %in% key_terms, !is.na(estimate)) %>%
  mutate(model_keep = dplyr::case_when(
    term %in% c("period_num","group_tNuevo_en_t") ~ "M4",
    term == "post_reform" ~ "M5",
    term == "tau" ~ "M6",
    TRUE ~ NA_character_
  )) %>%
  filter(model_std == model_keep)



  # coefs_key <- coefs_all %>%
  #   filter(term %in% key_terms) %>%
  #   filter(!is.na(estimate))

  if (nrow(coefs_key) == 0) {
    cat("**M5_3:** Coefficient files were read, but none contained the key terms: ",
        paste(key_terms, collapse = ", "), ".\n")
  } else {

    # --- 5) Compact TABLE: median + range across subsamples (avoids huge output)
    tab_diag_models <- coefs_key %>%
      group_by(family, pol_def, term) %>%
      summarise(
        n_cells = n(),
        est_med = median(estimate, na.rm = TRUE),
        est_min = min(estimate, na.rm = TRUE),
        est_max = max(estimate, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(
        value = paste0(fmt4(est_med), " [", fmt4(est_min), ", ", fmt4(est_max), "]")
      ) %>%
      select(family, pol_def, term, value, n_cells) %>%
      pivot_wider(names_from = term, values_from = value) %>%
      arrange(factor(family, levels = c("base_completa","subsamples_bill_origin","subsamples_vote_step","submuestras_balanceadas","other")),
              pol_def)

    # Ensure aligned colnames length = ncol
    cn <- names(tab_diag_models)
    align_vec <- rep("c", length(cn))

    kable(
      tab_diag_models,
      align = align_vec,
      caption = "Key coefficients across modules and polarization definitions. Entries report median estimate across available subsamples with range in brackets (min, max).",
      escape = FALSE
    )

# --- 6) FIGURE (cloud of points): all subsample estimates + jitter + color by IC95% ----

# (opcional) filtrar solo Model 4 si existe un identificador usable
coefs_plot <- coefs_key

# # Si en tus coeficientes hay una columna 'model' con algo tipo "Model 4" / "M4"
# if ("model" %in% names(coefs_plot)) {
#   coefs_plot <- coefs_plot %>%
#     mutate(model_chr = tolower(as.character(model))) %>%
#     filter(str_detect(model_chr, "model\\s*4\\b|\\bm4\\b"))
# }

# Si en vez de 'model' lo tienes en el nombre del archivo
# (descomenta si corresponde)
# coefs_plot <- coefs_plot %>%
#   filter(str_detect(tolower(basename(file)), "model4"))

coefs_plot <- coefs_plot %>%
  mutate(
    # diseño A–E
    design_code = case_when(
      family == "base_completa"           ~ "A",
      family == "subsamples_bill_origin"  ~ "B",
      family == "subsamples_vote_step"    ~ "C",
      family == "submuestras_balanceadas" ~ "D",
      TRUE                                ~ "E"
    ),
    design_code = factor(design_code, levels = c("A","B","C","D","E")),

    # etiquetas conceptuales (filas)
    term_lab = factor(
      term,
      levels = key_terms,
      labels = c("Trend", "Replacement", "Post-reform", "Δ since entry")
    ),

    # definiciones de polarización (columnas)
    pol_def_lab = factor(
      pol_def,
      levels = c("abs","sq_ideology")
    ),

    # color por IC95% (si hay std_error); si no, gris
    z = if_else(!is.na(std_error) & std_error > 0, estimate / std_error, NA_real_),
    sig_col = case_when(
      is.na(z)                ~ "SE not available",
      abs(z) > 1.96           ~ "IC95% excludes 0",
      TRUE                    ~ "IC95% crosses 0"
    ),
    sig_col = factor(sig_col, levels = c("IC95% crosses 0", "IC95% excludes 0", "SE not available"))
  ) %>%
  filter(is.finite(estimate), !is.na(term_lab), !is.na(pol_def_lab))

# Si quedó vacío (típico cuando el filtro "Model 4" no matchea), evita crash
if (nrow(coefs_plot) == 0) {
  cat("**M5_3:** No coefficients to plot after filtering (check how Model 4 is labeled in your coef files).\n")
} else {

  # Para superponer un resumen (mediana) por celda, construimos un df agregado
  med_df <- coefs_plot %>%
    group_by(design_code, pol_def_lab, term_lab) %>%
    summarise(est_med = median(estimate, na.rm = TRUE), .groups = "drop")

  p_diag <- ggplot(coefs_plot, aes(x = design_code, y = estimate)) +
    geom_hline(yintercept = 0, linewidth = 0.30) +

    # nube: puntos individuales con jitter horizontal
    geom_point(
      aes(color = sig_col),
      position = position_jitter(width = 0.18, height = 0, seed = 123),
      size = 0.75, alpha = 0.75
    ) +

    # mediana por celda (punto más grande encima)
    geom_point(
      data = med_df,
      aes(x = design_code, y = est_med),
      inherit.aes = FALSE,
      size = 1.35
    ) +

    scale_color_manual(
      values = c(
        "IC95% crosses 0"   = "black",
        "IC95% excludes 0"  = "red",
        "SE not available"  = "grey60"
      ),
      name = NULL
    ) +

    facet_grid(term_lab ~ pol_def_lab, scales = "free_y") +
    labs(x = NULL, y = "Coefficient estimate (points = subsamples; large dot = median)") +
    theme_minimal() +
    theme(
      panel.grid.minor = element_blank(),
      axis.text.x  = element_text(angle = 0, hjust = 0.5, size = 8),
      strip.text.x = element_text(size = 7),
      strip.text.y = element_text(size = 7),
      legend.position = "bottom",
      legend.text = element_text(size = 7),
      plot.margin  = margin(t = 5.5, r = 5.5, b = 10, l = 5.5)
    ) +
    labs(
      caption = paste0(
        "Design codes: A = Full sample (base_completa); B = Bill origin subsamples; ",
        "C = Vote-step subsamples; D = Balanced subsamples; E = Other.\n",
        "Color: red = |t|>1.96 using std_error (IC95% excludes 0); black = otherwise; grey = std_error unavailable."
      )
    ) +
    theme(plot.caption = element_text(hjust = 0, size = 7))


}

  }
}

# --- 6b) TABLE: share of subsample estimates with IC95% excluding 0 (red) ----

tab_sig_share <- coefs_plot %>%
  mutate(
    sig_bin = case_when(
      sig_col == "IC95% excludes 0" ~ "Excludes 0 (red)",
      sig_col == "IC95% crosses 0"  ~ "Crosses 0 (black)",
      TRUE                          ~ "SE not available (grey)"
    ),
    sig_bin = factor(sig_bin, levels = c("Excludes 0 (red)", "Crosses 0 (black)", "SE not available (grey)"))
  ) %>%
  group_by(term_lab, pol_def_lab, design_code, sig_bin) %>%
  summarise(n = n(), .groups = "drop_last") %>%
  mutate(
    n_total = sum(n),
    share   = n / n_total
  ) %>%
  ungroup()

# Compact wide table (shares, in %)
tab_sig_share_wide <- tab_sig_share %>%
  mutate(pct = 100 * share) %>%
  select(term_lab, pol_def_lab, design_code, sig_bin, pct, n, n_total) %>%
  tidyr::pivot_wider(
    names_from  = sig_bin,
    values_from = pct,
    values_fill = 0
  ) %>%
  arrange(term_lab, pol_def_lab, design_code)

fmt_pct <- function(x) ifelse(is.na(x), "", paste0(formatC(x, digits = 1, format = "f"), "%"))

tab_sig_share_wide_print <- tab_sig_share_wide %>%
  mutate(
    `Excludes 0 (red)`        = fmt_pct(`Excludes 0 (red)`),
    `Crosses 0 (black)`       = fmt_pct(`Crosses 0 (black)`),
    n_total = as.integer(n_total)
  ) %>%
  select(term_lab, pol_def_lab, design_code, n_total,
         `Excludes 0 (red)`, `Crosses 0 (black)`) %>%
  rename(
    Term   = term_lab,
    DV     = pol_def_lab,
    Design = design_code,
    N      = n_total
  )

knitr::kable(
  tab_sig_share_wide_print,
  align = c("l","l","c","c","c","c","c"),
  caption = "Share of subsample coefficient estimates whose IC95% excludes 0 (red) vs crosses 0 (black), by term, DV definition, and design."
)
```

```{r m5_3b, message=FALSE, warning=FALSE, echo=FALSE}
# print(p_diag)
```

#### Tail dynamics of ideological polarization

Figure S3 complements the main polarization analysis by focusing explicitly on the upper tail of the ideological distribution. While average dispersion captures broad shifts in legislative positioning, tail-based indicators help assess whether polarization is also driven by the expansion and concentration of ideological extremity among the most extreme legislators. All series are expressed as indices normalized to the first period (2002–06 = 1).

The first two panels track high-percentile absolute ideal points. Both p90 ($|x|$) and p95 ($|x|$) increase markedly over time, indicating that legislators at the upper end of the ideological spectrum become progressively more extreme. The rise in p90 ($|x|$) is steep through 2014–18 and remains elevated thereafter, while p95 ($|x|$) increases more gradually but reaches its highest level in 2022–26. These patterns show that the polarization trend documented in the main text is not confined to changes in average dispersion but is also reflected in the behavior of the most extreme legislators.

The third panel examines how much of quadratic dispersion is concentrated in the upper tail by reporting the share of total $\sum x^2$ accounted for by the top 10% most extreme legislators (ranked by $|x|$), again indexed to 2002–06. Values above one indicate a greater concentration of quadratic dispersion in the top decile relative to the baseline period. The series reveals non-monotonic dynamics: concentration rises in mid-periods, drops sharply in 2018–22, and rebounds strongly in 2022–26. This pattern suggests an important nuance. Even as high-percentile extremity increases, the extent to which overall polarization is disproportionately driven by a small subset of extreme legislators varies across periods. Substantively, this is consistent with shifts in composition and with changes in the balance between broad-based dispersion and tail-driven polarization.

```{r m5_4, message=FALSE, warning=FALSE, echo=FALSE, fig.cap="Upper tails of the ideological distributions", fig.pos='H'}

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)
  library(here)
})

aqui <- here()

# --- 1) Locate baseline ideal-points file (point estimates) -------------------
cands <- c(
  file.path(aqui, "pape/polarizacion/data/base_completa/ideologia_diputados_largo_emIRT_con_anclas.csv"),
  file.path(aqui, "pape/polarizacion/data/ideologia_diputados_largo_emIRT_con_anclas.csv")
)
path_ip <- cands[file.exists(cands)][1]
stopifnot(!is.na(path_ip), file.exists(path_ip))

ip <- read_csv(path_ip, show_col_types = FALSE)

pick_first <- function(df, pattern) {
  nm <- names(df)
  hit <- nm[str_detect(tolower(nm), pattern)]
  if (length(hit) == 0) NA_character_ else hit[1]
}

# --- 2) Robust column detection -----------------------------------------------
col_id <- pick_first(ip, "^diputadoid$|^nombreid$|diputado_id|nombre_id|^id$")
col_pn <- pick_first(ip, "period_num|^period$|^periodo$|term|session")
col_pr <- pick_first(ip, "^periodo$|periodo_|period_label|period_lab")
col_x  <- pick_first(ip, "^ideologia$|^ideology$|ideal(?!.*low)(?!.*high)")

stopifnot(!is.na(col_x))  # we need the point estimate
if (is.na(col_id))  col_id <- pick_first(ip, "diputado|nombre")  # very last-resort
stopifnot(!is.na(col_id))

ip2 <- ip %>%
  transmute(
    id        = as.character(.data[[col_id]]),
    period_num_raw = if (!is.na(col_pn)) suppressWarnings(as.integer(.data[[col_pn]])) else NA_integer_,
    periodo_raw    = if (!is.na(col_pr)) as.character(.data[[col_pr]]) else NA_character_,
    x         = suppressWarnings(as.numeric(.data[[col_x]]))
  ) %>%
  mutate(
    period_num = if_else(
      !is.na(period_num_raw),
      period_num_raw,
      suppressWarnings(as.integer(str_extract(periodo_raw, "\\d+")))
    )
  ) %>%
  select(id, period_num, x) %>%
  filter(!is.na(period_num), !is.na(x)) %>%
  filter(is.finite(x))

# Period labels (edit if needed)
period_labels <- c("1"="2002–06","2"="2006–10","3"="2010–14","4"="2014–18","5"="2018–22","6"="2022–26")
ip2 <- ip2 %>%
  mutate(period_lab = recode(as.character(period_num), !!!period_labels))

# --- 3) Tail diagnostics per period -------------------------------------------
diag <- ip2 %>%
  group_by(period_num, period_lab) %>%
  summarise(
    n = n(),
    p90_abs = as.numeric(quantile(abs(x), 0.90, na.rm = TRUE, names = FALSE)),
    p95_abs = as.numeric(quantile(abs(x), 0.95, na.rm = TRUE, names = FALSE)),
    share_top10_sq = {
      thr <- as.numeric(quantile(abs(x), 0.90, na.rm = TRUE, names = FALSE))
      top <- abs(x) >= thr
      denom <- sum(x^2, na.rm = TRUE)
      if (is.finite(denom) && denom > 0) sum((x^2)[top], na.rm = TRUE) / denom else NA_real_
    },
    .groups = "drop"
  ) %>%
  arrange(period_num)

# Index to first period (2002–06 = 1)
base_p <- min(diag$period_num, na.rm = TRUE)
diag_idx <- diag %>%
  mutate(
    p90_abs_idx       = p90_abs / p90_abs[period_num == base_p],
    p95_abs_idx       = p95_abs / p95_abs[period_num == base_p],
    share_top10_sq_idx = share_top10_sq / share_top10_sq[period_num == base_p]
  )

plot_df <- diag_idx %>%
  select(period_num, period_lab,
         `p90(|x|)` = p90_abs_idx,
         `p95(|x|)` = p95_abs_idx,
         `Share of Σx² from top 10% |x|` = share_top10_sq_idx) %>%
  pivot_longer(-c(period_num, period_lab), names_to = "metric", values_to = "value")

p_tail <- ggplot(plot_df, aes(x = period_num, y = value)) +
  geom_line(linewidth = 0.45, aes(group = metric)) +
  geom_point(size = 1.0) +
  facet_wrap(~ metric, ncol = 3, scales = "free_y") +
  scale_x_continuous(
    breaks = sort(unique(plot_df$period_num)),
    labels = plot_df %>%
      distinct(period_num, period_lab) %>%
      arrange(period_num) %>% pull(period_lab)
  ) +
  labs(x = NULL, y = "Index (2002–06 = 1)") +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1))

p_tail

```

## References
